Using reference models in variable selection \vspace{.1in}

### 2020-04-26 paper/ref_approach.tex
improved tables, captions and float placements

The inclusion frequencies of each variable in the final model given 100
bootstrap samples are shown in Figure
\ref{fig:inclusion_frequencies}. In case of projpred there are two
variables, `abdomen' and `weight', which have inclusion frequencies
above 50\% (`abdomen' is the only one included always), the third most
frequently included is `wrist' at 44\%, and the fourth one is `height'
at 35\%.  The steplm approach has seven variables with inclusion
frequencies above 50\%. Such a higher variability and lower stability
of steplm can be observed also in the bootstrap model selection
frequencies reported in Table \ref{tab:model_frequencies}. For
example, the first five selected models have a cumulative frequency of
76\% with projpred, but only of 14\% with steplm. In addition, the
sizes of the selected models with projpred are much smaller than the ones 
selected with steplm.
\begin{figure}[tp]
  \centering
  \includegraphics[width=0.98\textwidth]{graphics/inc_prob.pdf}
  \caption{Bootstrap inclusion frequencies after 100 bootstrap samples.}
  \label{fig:inclusion_frequencies}
\end{figure}

\begin{table}[tp]
\footnotesize
\centering
\begin{tabular}{l|lr|lr}
%  \hline
M & projpred & Freq \% & steplm & Freq \%  \\ 
  \hline
1 & abdom., weight & 39 & abdom., age, forearm, height, hip, neck, thigh, wrist & 4 \\
2 & abdom., wrist & 10 & abdom., age, chest, forearm, height, neck, thigh, wrist & 4 \\
3 & abdom., height & 10 & abdom., forearm, height, neck, wrist & 2 \\
4 & abdom., height, wrist & 9 & abdom., forearm, neck, weight, wrist & 2 \\
5 & abdom., weight, wrist & 8 & abdom., age, height, hip, thigh, wrist & 2 \\
6 & abdom., chest, height, wrist & 2 & abdom., age, height, hip, neck, thigh, wrist & 2 \\
7 & abdom., biceps, weight, wrist & 2 & abdom., age, ankle, forearm, height, hip, neck, thigh, wrist & 2 \\
8 & abdom., height, weight, wrist & 2 & abdom., age, biceps, chest, height, neck, wrist & 2 \\
9 & abdom., age, wrist & 2 & abdom., age, biceps, chest, forearm, height, neck, thigh, wrist & 2 \\
10 & abdom., age, height, neck, thigh, wrist & 2 & abdom., age, ankle, biceps, weight, wrist & 2 \\
%   \hline
\end{tabular}
\caption{Bootstrap model selection frequencies after 100 bootstrap samples.}
\label{tab:model_frequencies}
\end{table}

The first two rows of Table \ref{tab:model_performances} show the
predictive performances, in terms of cross-validated root mean square
error (RMSE), of the full model and the selected models using projpred
or steplm.  There is no significant difference in predictive performance
of the selected models by different approaches even if there is
difference in the selected variables. This can be explained by high
correlation between the variables and different combinations can
provide similar predictive accuracy.

@@

\begin{table}[tp]
\footnotesize
\centering
\begin{tabular}{ll|rrrr}
%  \hline
Data  & Method & RMSE Full & RMSE Sel & \# Sel (CPL, AVG, SD) & \# Sel noisy (CPL, AVG, SD) \\ 
  \hline
body fat & projpred & 4.36 & 4.45 & (2, 2.3, 0.48) &   \\
& steplm & 4.38 & 4.46 & (7, 6, 0.94) &  \\
\hline
+ noisy variables & projpred & 4.48 & 4.46 & (2, 2, 0)&  (0, 0, 0)  \\
& steplm & 5.74 & 5.13 & (23, 26.2, 4.54) & (15, 18.9, 4.01) \\
%   \hline
\end{tabular}
\caption{Predictive performances with original body fat data (first two
  rows) and with extra noisy variables (last two rows) estimated with
  10-fold cross-validation. Abbreviations: RMSE = root mean squared error;
  Full = full model; Sel = selected submodel; 
  \# Sel = total number of selected variables;
  \# Sel noisy = number of selected noisy variables. 
  CPL = results for the complete data (not using cross-validation).
  AVG = average over the 10-fold cross-validation. SD = standard deviation
  over the 10-fold cross-validation.}
\label{tab:model_performances}
\end{table}

@@

When using a reference model, we first obtain predictions for $f$
using all the variables $\{X_{j}\}_{j=1}^{p}$ taking into account
that we have only observed the noisy representation $Y$ of $f$. If our
model is good, we are able to describe $f$ better than $Y$ itself can,
which improves the accuracy in the estimation of the relevance of the
variables.  Figure \ref{fig:correlation} illustrates this process in
the form of a scatter plot of (absolute) correlations of the
variables with $Y$ against the corresponding correlations with the
predictions of a reference model (in this case the posterior
predictive means of model \eqref{eq:ref_mod}; see Section
\ref{simulations}). Looking at the marginal distributions, we see
that using a reference model to filter out noise in the data, the two
groups of variables (relevant and non-relevant) can be distinguished
much better than when the correlation is computed using the observed
noisy data directly.

\begin{figure}[tp]
  \centering
  \vspace{-5mm}
  \includegraphics[width=0.4\textwidth]{graphics/correlation.pdf}
  \caption{Sample correlation plot of each variable (relevant in red, 
  non-relevant in blue) with the target variable $y$ and the latent variable 
  $f$ respectively on the x and y axis. Data are generated according to 
  \eqref{eq:simulated_data} with parameters $n=70$, $\rho=0.3$,
  $k=100$, $p=1000$.\\}
  \label{fig:correlation}
\end{figure}

@@

\subsection{Simulation study}

@@

When the variable selection is repeated with different simulated data
sets, there is some variability in the selected variables. We
measure the stability of variable selection by computing the entropy
of the observed distribution of the included variables over different models. The smallest entropy
would be obtained if the approach always selected the same set of
variables, and the largest entropy would be observed if the approach
would always selected different sets of variables. Therefore, lower entropy 
corresponds to a more stable selection. Highly correlated predictive variables may happen to be
selected alternately, thus making stability estimation of the
selection a non-trivial task. Entropy can not distinguish the interchangeability 
due to correlation from instability. Thus, such a measure should be 
considered as a relative, and not as an absolute, measure of stability.
Figure \ref{fig:entropy} shows the entropy scores for the different compared
methods. The use of a reference model
improves the stability of steplm in variable selection slightly, while it makes
little difference for the Bayesian linear regression. The
projpred approach turns out to be far more stable than all
other methods. This is likely due to projpred being based on better
decision theoretical formulation which 1) takes into account the full
predictive distribution and not just point estimate and 2) projects
the reference model posterior to the submodel instead of using a
simple refit of submodels.

% TODO: Bigger and more distinctive symbols in figures
% label order: steplm, bayes.step, projpred
\begin{figure}[tp]
  \centering
  \includegraphics[width=0.98\textwidth]{graphics/rmse_vs_fdr_parallel.pdf}
  \caption{Root mean square error (RMSE) against false discovery rate in the minimal subset variable selection with one standard deviation error bars.}
  \label{fig:rmse_vs_fdr}
\end{figure}

% TODO: Bigger and more distinctive symbols in figures
% TODO: order steplm, bayes.step, projpred
\begin{figure}[tp]
  \centering
  \includegraphics[width=0.98\textwidth]{graphics/entropy_parallel.pdf}
  \caption{Entropy score in the minimal subset variable selection.}
  \label{fig:entropy}
\end{figure}

@@

Here we repeat the selection of Section \ref{bodyfat-1} via
stepwise backward regression. In this case, the overall number of
variables (original plus noisy) is 100, as it was in the last part of
Section \ref{bodyfat-1}. We compare results with and without using
a simple reference model approach outlined in \eqref{eq:ref_mod} with
steplm. Figure \ref{fig:bodyfat_step_refvsdata} shows the number of
irrelevant variables included in the final model and the out-of-sample
root mean square error (RMSE). Results are based on 100 bootstrap
samples on the whole dataset, and the predictive performance is tested
on the observations excluded at each bootstrap sample. We observe that
the reference model reduces the number of irrelevant variables included
in the final model. This leads to less overfitting and thus to
improved out-of-sample predictive performance in terms of RMSE. The
reference model approach applied to the stepwise backward regression
achieves outstanding improvements considering its simplicity, yet it
does not reach the goodness of the much more sophisticated projective
prediction approach (see results of Section \ref{bodyfat-1}).
\begin{figure}[tp]
  \centering
  \includegraphics[width=0.98\textwidth]{graphics/bodyfat_step_refvsdata.pdf}
  \caption{Stepwise backward selection with and without using a reference model. The x-axis denotes the number of selected irrelevant variables on the left and the out-of-sample RMSE on the right-hand side based on 100 bootstrap samples.\\}
  \label{fig:bodyfat_step_refvsdata}
\end{figure}

\begin{table}[tp]
\footnotesize
\centering
\begin{tabular}{l|r|r}
%  \hline
Method & RMSE & Selected noisy \\ 
  \hline
Stepwise selection & 6.92 (0.68) & 48.1 (7.46)    \\
Reference model + Stepwise selection & 4.71 (0.45) & 25.1 (6.89)   \\
\end{tabular}
\caption{Mean and standard deviation (between brackets) values of Figure \ref{fig:bodyfat_step_refvsdata}}
\label{tab:step_refvsdata}
\end{table}

@@

\subsection{Simulation study}
\label{simulations}

@@

% \begin{figure}[tp]
%   \centering
%   \includegraphics[width=0.98\textwidth]{graphics/k_RHS.pdf}
%   \caption{Average posterior mean values for the shrinkage factors
%     using the regularised horseshoe prior based on 100
%     simulations. Relevant and non-relevant variables are displayed in
%     red and blue respectively.\\}
%   \label{fig:k}
% \end{figure}

@@

\begin{figure}[tp]
  \centering
  \includegraphics[width=0.98\textwidth]{graphics/sensitivity_vs_fdr_iterated.pdf}
  \caption{Sensitivity against false discovery rate based on 100 data simulations with one standard deviation error bars.\\}
  \label{fig:sensitivity_vs_fdr_iterated}
\end{figure}

@@

\begin{figure}[tp]
  \centering
  \includegraphics[width=0.98\textwidth]{graphics/stability_iterated.pdf}
  \caption{Stability estimates with 95\% intervals based on 100 data simulations.\\}
  \label{fig:stability_iterated}
\end{figure}

@@

@@

\begin{figure}[tp]
  \centering
  \includegraphics[width=0.98\textwidth]{graphics/bodyfat_sensitivity_vs_fdr.pdf}
  \caption{Sensitivity against false discovery rate based on 100 bootstrap samples with one standard deviation error bars.\\}
  \label{fig:bodyfat_sensitivity_vs_fdr}
\end{figure}

\begin{figure}[tp]
  \centering
  \includegraphics[width=0.98\textwidth]{graphics/bodyfat_stability.pdf}
  \caption{Stability estimates with 0.95 confidence intervals based on 100 bootstrap samples.\\}
  \label{fig:bodyfat_stability}
\end{figure}

>>>>

The inclusion frequencies of each variable in the final model given 100
bootstrap samples are shown in Figure
\ref{fig:inclusion_frequencies}. In case of projpred there are two
variables, `abdomen' and `weight', which have inclusion frequencies
above 50\% (`abdomen' is the only one included always), the third most
frequently included is `wrist' at 44\%, and the fourth one is `height'
at 35\%.  The steplm approach has seven variables with inclusion
frequencies above 50\%. Such a higher variability and lower stability
of steplm can be observed also in the bootstrap model selection
frequencies reported in Table \ref{tab:model_frequencies}. For
example, the first five selected models have a cumulative frequency of
76\% with projpred, but only of 14\% with steplm. In addition, the
sizes of the selected models with projpred are much smaller than the ones 
selected with steplm.
\begin{figure}[tp]
  \centering
  \includegraphics[width=0.98\textwidth]{graphics/inc_prob.pdf}
  \vspace{-0.7\baselineskip}
  \caption{Body fat example: Bootstrap inclusion frequencies calculated from 100 bootstrap samples. The projpred approach has less variability on which variables are selected.}
  \label{fig:inclusion_frequencies}
\end{figure}

\begin{table}[tp]
\footnotesize
\centering
\begin{tabular}{l|lr|lr}
%  \hline
M & projpred & Freq \% & steplm & Freq \%  \\ 
  \hline
1 & abdom., weight & 39 & abdom., age, forearm, height, hip, neck, thigh, wrist & 4 \\
2 & abdom., wrist & 10 & abdom., age, chest, forearm, height, neck, thigh, wrist & 4 \\
3 & abdom., height & 10 & abdom., forearm, height, neck, wrist & 2 \\
4 & abdom., height, wrist & 9 & abdom., forearm, neck, weight, wrist & 2 \\
5 & abdom., weight, wrist & 8 & abdom., age, height, hip, thigh, wrist & 2 \\
6 & abdom., chest, height, wrist & 2 & abdom., age, height, hip, neck, thigh, wrist & 2 \\
7 & abdom., biceps, weight, wrist & 2 & abdom., age, ankle, forearm, height, hip, neck, thigh, wrist & 2 \\
8 & abdom., height, weight, wrist & 2 & abdom., age, biceps, chest, height, neck, wrist & 2 \\
9 & abdom., age, wrist & 2 & abdom., age, biceps, chest, forearm, height, neck, thigh, wrist & 2 \\
10 & abdom., age, height, neck, thigh, wrist & 2 & abdom., age, ankle, biceps, weight, wrist & 2 \\
%   \hline
\end{tabular}
\caption{Body fat example: Bootstrap model selection frequencies from 100 bootstrap samples.  The projpred approach has less variability on which variable combinations are selected.}
\label{tab:model_frequencies}
\end{table}

The first two rows of Table \ref{tab:model_performances} show the
predictive performances, in terms of cross-validated root mean square
error (RMSE), of the full model and the selected models using projpred
or steplm.  There is no significant difference in predictive performance
of the selected models by different approaches even if there is
clear difference in the number of selected variables. This can be explained by high
correlation between the variables and different combinations can
provide similar predictive accuracy.

@@

\begin{table}[tp]
\footnotesize
\centering
\begin{tabular}{ll|rrrr}
%  \hline
Data  & Method & RMSE Full & RMSE Sel & \# Sel (10-CV avg, sd) & \# Sel noisy (10-CV avg, sd) \\ 
  \hline
body fat & projpred & 4.4 & 4.5 &  2\, (2.3, 0.5) &   \\
& steplm & 4.4 & 4.5 & 7\, (6.0, 0.9) &  \\
\hline
+ noisy variables & projpred & 4.5 & 4.5 &  2\, (2.0, 0.0)&  0\,\; (0,\, 0)  \\
& steplm & 5.7 & 5.1 & 23\, (26,\; 4.5) &  15\, (19, 4) \\
%   \hline
\end{tabular}
\caption{Body fat example: Predictive performances with original data (first two
  rows) and with extra noisy variables (last two rows) estimated with
  10-fold cross-validation. Abbreviations: RMSE = root mean squared error;
  Full = full model; Sel = selected submodel; 
  \# Sel = total number of selected variables;
  \# Sel noisy = number of selected noisy variables. 
  10-CV avg = average over the 10 folds in cross-validation. 10-CV sd = standard deviation
  over the 10 folds in cross-validation.}
\label{tab:model_performances}
\end{table}

@@

When using a reference model, we first obtain predictions for $f$
using all the variables $\{X_{j}\}_{j=1}^{p}$ taking into account
that we have only observed the noisy representation $Y$ of $f$. If our
model is good, we are able to describe $f$ better than $Y$ itself can,
which improves the accuracy in the estimation of the relevance of the
variables.  Figure \ref{fig:correlation} illustrates this process in
the form of a scatter plot of (absolute) correlations of the
variables with $Y$ against the corresponding correlations with the
predictions of a reference model (in this case the posterior
predictive means of model \eqref{eq:ref_mod}; see Section
\ref{simulations}). Looking at the marginal distributions, we see
that using a reference model to filter out noise in the data, the two
groups of variables (relevant and non-relevant) can be distinguished
much better than when the correlation is computed using the observed
noisy data directly.
\begin{figure}[tp]
  \centering
  \vspace{-5mm}
  \includegraphics[width=0.35\textwidth]{graphics/correlation.pdf}
  \vspace{-0.5\baselineskip}
  \caption{Sample correlation plot of each variable (relevant in red, 
  non-relevant in blue) with the target variable $y$ and the latent variable 
  $f$ respectively on the x and y axis. Simulated data are generated according to 
  \eqref{eq:simulated_data} with parameters $n=70$, $\rho=0.3$,
  $k=100$, $p=1000$.}
  \label{fig:correlation}
\end{figure}

@@

\subsection{Simulation study 1}

@@

When the variable selection is repeated with different simulated data
sets, there is some variability in the selected variables. We
measure the stability of variable selection by computing the entropy
of the observed distribution of the included variables over different models. The smallest entropy
would be obtained if the approach always selected the same set of
variables, and the largest entropy would be observed if the approach
would always selected different sets of variables. Therefore, lower entropy 
corresponds to a more stable selection. Highly correlated predictive variables may happen to be
selected alternately, thus making stability estimation of the
selection a non-trivial task. Entropy can not distinguish the interchangeability 
due to correlation from instability. Thus, such a measure should be 
considered as a relative, and not as an absolute, measure of stability.
Figure \ref{fig:entropy} shows the entropy scores for the different compared
methods. The use of a reference model
improves the stability of steplm in variable selection slightly, while it makes
little difference for the Bayesian linear regression. The
projpred approach turns out to be far more stable than all
other methods. This is likely due to projpred being based on better
decision theoretical formulation which 1) takes into account the full
predictive distribution and not just point estimate and 2) projects
the reference model posterior to the submodel instead of using a
simple refit of submodels.
% TODO: Bigger and more distinctive symbols in figures
% label order: steplm, bayes.step, projpred
\begin{figure}[tp]
  \centering
  \includegraphics[width=0.98\textwidth]{graphics/rmse_vs_fdr_parallel.pdf}
  \vspace{-0.3\baselineskip}
  \caption{Simulation study 1: Root mean square error (RMSE) against false discovery rate in the minimal subset variable selection with one standard deviation error bars. The projpred approach has the smallest RMSE and false discovery ratio.}
  \label{fig:rmse_vs_fdr}
\end{figure}
% TODO: Bigger and more distinctive symbols in figures
% TODO: order steplm, bayes.step, projpred
\begin{figure}[tp]
  \centering
  \includegraphics[width=0.98\textwidth]{graphics/entropy_parallel.pdf}
  \vspace{-0.3\baselineskip}
  \caption{Simulation study 1: Entropy score in the minimal subset variable selection.  The projpred approach has much smaller entropy score than the other approaches.}
  \label{fig:entropy}
\end{figure}

@@

Here we repeat the selection of Section \ref{bodyfat-1} via
stepwise backward regression. In this case, the overall number of
variables (original plus noisy) is 100, as it was in the last part of
Section \ref{bodyfat-1}. We compare results with and without using
a simple reference model approach outlined in \eqref{eq:ref_mod} with
steplm. Figure \ref{fig:bodyfat_step_refvsdata} shows the number of
irrelevant variables included in the final model and the out-of-sample
root mean square error (RMSE). Results are based on 100 bootstrap
samples on the whole dataset, and the predictive performance is tested
on the observations excluded at each bootstrap sample. We observe that
the reference model reduces the number of irrelevant variables included
in the final model. This leads to less overfitting and thus to
improved out-of-sample predictive performance in terms of RMSE. The
reference model approach applied to the stepwise backward regression
achieves outstanding improvements considering its simplicity, yet it
does not reach the goodness of the much more sophisticated projective
prediction approach (see results of Section \ref{bodyfat-1}).
\begin{figure}[tp]
  \centering
  \includegraphics[width=0.98\textwidth]{graphics/bodyfat_step_refvsdata.pdf}
  \vspace{-0.7\baselineskip}
  \caption{Body fat example: Stepwise backward selection with and without using a reference model. The x-axis denotes the number of selected irrelevant variables on the left and the out-of-sample RMSE on the right-hand side based on 100 bootstrap samples. The reference approach reduces the number of noisy variables selected and the out-of-sample RMSE.}
  \label{fig:bodyfat_step_refvsdata}
\end{figure}
\begin{table}[tp]
\footnotesize
\centering
\begin{tabular}{l|rr}
%  \hline
Method & RMSE & Selected noisy \\ 
  \hline
Stepwise selection & 6.9 (0.7) & 48 (7)    \\
Reference model + Stepwise selection & 4.7 (0.5) & 25 (7)   \\
\end{tabular}
\caption{Body fat example: Mean and standard deviation (between brackets) values of Figure \ref{fig:bodyfat_step_refvsdata}.}
\label{tab:step_refvsdata}
\end{table}

@@

\subsection{Simulation study 2}
\label{simulations}

@@

% \begin{figure}[tp]
%   \centering
%   \includegraphics[width=0.98\textwidth]{graphics/k_RHS.pdf}
%   \caption{Average posterior mean values for the shrinkage factors
%     using the regularised horseshoe prior based on 100
%     simulations. Relevant and non-relevant variables are displayed in
%     red and blue respectively.\\}
%   \label{fig:k}
% \end{figure}
\begin{figure}[tp]
  \centering
  \includegraphics[width=0.98\textwidth]{graphics/sensitivity_vs_fdr_iterated.pdf}
  \vspace{-0.3\baselineskip}
  \caption{Simulation study 2: Complete variable selection sensitivity against false discovery rate based on 100 data simulations with one standard deviation error bars. The reference approach improves sensitivity and reduces false discovery rate for all methods.}
  \label{fig:sensitivity_vs_fdr_iterated}
\end{figure}
\begin{figure}[tp]
  \centering
  \includegraphics[width=0.98\textwidth]{graphics/stability_iterated.pdf}
  \vspace{-0.3\baselineskip}
  \caption{Simulation study 2: Complete variable selection stability estimates with 95\% intervals based on 100 data simulations. The reference approach improves stability for all methods.}
  \label{fig:stability_iterated}
\end{figure}

@@

@@

@@

\begin{figure}[tp]
  \centering
  \includegraphics[width=0.98\textwidth]{graphics/bodyfat_sensitivity_vs_fdr.pdf}
  \vspace{-0.3\baselineskip}
  \caption{Body fat example with noisy variables: Complete variable selection sensitivity against false discovery rate based on 100 bootstrap samples with one standard deviation error bars. The improvement from using the reference approach is small (except that the projpred is much better than lasso).}
  \label{fig:bodyfat_sensitivity_vs_fdr}
\end{figure}
\begin{figure}[tp]
  \centering
  \includegraphics[width=0.98\textwidth]{graphics/bodyfat_stability.pdf}
  \vspace{-0.3\baselineskip}
  \caption{Body fat example with noisy variables: Complete variable selection stability estimates with 0.95 confidence intervals based on 100 bootstrap samples. The improvement from using the reference approach is small (except that the projpred is much better than lasso).}
  \label{fig:bodyfat_stability}
\end{figure}

@@
